---
title: "Intro to R: Session 1"
output:
  html_notebook:
  pdf_document:
  html_document:
editor_options: 
  markdown: 
    wrap: sentence
---

In an R Markdown document, you can type and **format** text, and also include images.
You can execute code in the console or in code chunks, as the one here below.
To run a line of code, put the cursor on the same line then press **ctrl + enter**.
To run the entire code chunk press **ctrl + shift + enter**.

### First chunk

```{r first chunk}

1 + 1 

#This is a comment

2 + 2 #this is also a comment

#you can comment or un-comment a line by highlighting it 
#then pressing ctrl + shift + c

"this is not a comment"


```

### R as calculator

```{r}

5 + 2 #Addition 
5 - 2 #Subtraction
5 * 2 #Multiplication
5 / 2 #Division
5^2   #Exponent
1+3*5 #mixed
(1+2)+2^2*3 #can you guess the answer?

```

But R is more than just a calculator.
You can carry out more complex operations using *functions*.

### Functions

```{r}

#get the square root of 4
4^(0.5)

#or using the function: sqrt()
sqrt(4)

#get the sum of 2 + 3
2 + 3

#or using the function: sum()
sum(2,3)

#find max value in a series of numbers 
max(1,2,3,4)

#or minimum
min(1,2,3,4)

#round a number
round(3.142)

#get the absolute value of a number: |x|
abs(-3)
abs(3)

#you can print a text
print("Functions make life easier") 

```

Feel free to do some operations and use some functions in the following chunk and in the console too.
To clear the console, click at a new line on the console then press ctrl + l.

```{r}






```

So the form of a function is the name of the function followed by ( ).
What goes inside the parentheses is called the *arguments* of the function.
So far, we only provided the function with an input argument, but further arguments can be provided as extra options for how the function would work.

### Functions with arguments

```{r}

#remember the rounding task?
round(3.142)

#it is equivalent to
round(3.142, digits = 0)


#what if we want rounding to the 1 or 2 decimal place?
round(3.142, digits = 1)
round(3.142, digits = 2)


#if you are more familiar with the arguments of the function
#you can skip providing the name of the argument
round(3.142, 1)
round(3.142, digits = 1)
round(x = 3.142, digits = 1)


#to get more information about a function and its arguments
#type ? then the name of the function

?round

#functions can take other functions as input
sqrt(2)           #answer is approx. 1.414214
round(sqrt(2), 1) #answer is 1.4

#can make it even more complicated, r works from inside out
round(sqrt(sum(1,1)), 1)

#notice where every function starts and ends
#by clicking after a bracket to see the bracket closing it 
round(   
  sqrt(
    sum(1,1)
             )
              , 1)

#did you notice that r does not care about spaces and new lines?


#lets look at the paste() function
paste("Welcome", "to", "Mannheim")
paste("Welcome", "to", "Mannheim", sep = " ")

#what is the time?
paste("11", "45", sep = ":")

#what happens if you remove the "sep ="?
paste("11", "45", ":")


#let's add seconds
paste("11", "45", "12", sep = ":")


#be creative
paste("Peter Parker", "MJ", sep = " *hearts* ")

```

lets take some time to play with functions.

### Some Exercises

**Problem 1**

You want to start an online platform to fight misinformation.
You expect the project to start financing itself after **10** years, so you get a loan from the bank for **1.9** million euros with an interest rate of **2.5%** for the entire period.

How much do you expect to have paid the bank by the end of the period?

```{r}

#you can use this following formula:
#total amount = the principal * (1 + interest rate * years)




#round the amount to the first and second decimal place



#hint: first answer is 2.37 millions, second is 2.4 millions
```

**Problem 2**

A voter has to choose between two candidates.
She has a position of **5** on the cultural left-right scale and **2** on the economic left-right scale.
She puts a weight of **0.7** on cultural issues and **0.3** on economic issues.
Candidate A has position of **6** on both cultural and economic issues, while candidate B has position of **3** on both cultural and economic issues.
Which candidate is she more likely to vote for?

```{r}

#hint: she is likely to vote for the candidate with the smaller distance
#to her issue positions, use these formulas:
# D_A = W_E * |P_EV - P_EA| + W_C *|P_CV - P_CA|
# D_B = W_E * |P_EV - P_EB| + W_C *|P_CV - P_CB|

#where:
#D_i is distance of voter to Candidate i
#W_E is the weight the voter puts on economic issues
#W_C is the weight the voter puts on cultural issues
#P_Ei is the position of person i on the economic scale
#P_Ci is the position of person i on the cultural scale










###hint: D_A = 1.9, D_B = 1.7

```

As you must have noticed in the last example.
The formula for choosing the smallest distance could get quite complicated.

```{r}

min(0.3 * abs(2-6) + 0.7 *abs(5-6), 0.3 * abs(2-3) + 0.7 *abs(5-3))

#this would have been easier

D_A <- 0.3 * abs(2 - 6) + 0.7 *abs(5 - 6)
D_B <- 0.3 * abs(2 - 3) + 0.7 *abs(5 - 3)

min_D <- min(D_A, D_B)


```

What did we do here?
We stored the distances in an **object**.
This is a key advantage of using R (and other programming languages), the ability to store the results of your calculations for further manipulations.
You can think of objects as storage containers for your calculations.

### Objects

```{r}

#you can assign an object using the <- operator

age <- 23
income <- 1400
field <- "political science"
university <- "University of Mannheim"


#you can also use the = operator but it is not recommended
age = 23

#stored objects appear in the environment
#to free up memory, you can remove objects using rm() command
rm(age)

#you can remove all objects using the command
rm(list = ls())


```

There are rules for naming objects.

-   Objects can contain numbers but must start with a character: **2014GDP** is invalid while **GDP2014** is valid.

-   Object names should not contain a blank space, use an underscore "**\_**" instead.
    For example: **vote_share** instead of **vote share**.

-   Avoid using other special characters such as **? ! \*** and so on.

-   Object names in R are case sensitive, so object *Age* is different from *age*.

-   Avoid using names of functions such as sum, mean, and print...etc for your objects.

    A good coding practice is not to **hard code** your values.
    lets look at an example in the following chunk:

```{r}

#if you are calculating the area of a trianlge with base length 2 cm and height 3 cm, you can go about it like this:

tri_area <- 0.5 * 2 * 3

#but a "cleaner" way of doing it would be:
base <- 2
height <- 3
tri_area <- 0.5 * base * height

#this way you don't have to manually change the "2" and "3" everywhere if you need to use these dimensions again later in the code


#let's repeat the example with the loan from the bank for the anti-misinformation platform, this time without hard coding below is the text of the problem again:

#You want to start an online platform for fight misinformation
#you expect the project to start financing itself after 10 years
#you get a loan from the bank for 1.9 million euros
#the interest rate is about 2.5% for the entire period
#how much do you expect to have paid the bank by the end of the period
#you can use this following formula:
# total amount = the principal * (1 + interest rate * years)






```

Objects can store different *data types*:

1.  Numeric, basically numbers such as 0, 1, 1.2, -32.1, and -3.21e1
2.  Character, text such as "vote", "SPD", "1". Character input should always be provided between single ' ' or double " " quotation marks, otherwise R will think your input is referring to the name of an object that does not exist.
3.  Logical, which is true or false statement. In the syntax of R: **TRUE** or **T** and **FALSE** or **F**.
4.  Factors, which is similar to character but is used to indicate a category, such as "Republican", "Democrat", "Independent".
5.  There are other data types that are less frequently used, such as a Date format, but we will focus on the main ones here.

You can check the type of an object using the `class()` command, you can also convert the type of an object (where possible) with commands such as `as.numeric` .
Let's have a look below:

```{r}

area <- 15.1

gdp_growth  <- -1.2

country <- "DE"

grade <- "1.3"

Mannheim_is_fun <- TRUE #notice the color of TRUE

putin_will_win <- F  #F or T as shorthand for FALSE or TRUE


#Let's check the classes 

class(gdp_growth)
class(country)
class(putin_will_win)


#let's convert the class of grade into numeric
class(grade)

grade <- as.numeric(grade)

class(grade)

#you can convert grade into a character with as.character()
grade <- as.character(grade)

#expect an error message if you try to convert country into numeric


#TRUE and FALSE are treated like 1 and 0
number <- 1
number <- number + TRUE
number <- number * FALSE

```

If some data is not acting as you are expecting, there is a chance the type is not what you want it to be, e.g., you are processing the data as numeric but what you have is character or factor.
Just keep an eye on the class of your data.

Very well, now take few minutes to create your own objects and assign values to them.
Check their classes and try to convert between data types.
See what works and what does not.

```{r}









```

### Vectors

Almost always we don't work with one single data point, but our dataset has many observations.
For example, the age of one person is not enough to establish the relationship between one's age and vote preferences.
We need a series of age observations.
A series of observation in R is called a **vector**.
You can use the command `c()` to create a vector from your input.

```{r}

philosophers <- c("Hegel", "Nietzsche", "Leibniz", "kant")

print(philosophers)

age <- c(23, 20, 22, 24, 21, 26, 19)

print(age)

revolution_success <- c(TRUE, FALSE, FALSE, TRUE, TRUE)

print(revolution_success)

#you can the number of elements in a vector using length()
n_age <- length(age)

#we can use our functions on vectors
mean_age <- sum(age)/length(age)

#or easier
mean_age <- mean(age)


#same as before with min and max
min_age <- min(age)
max_age <- max(age)


#you can perform mathematical operations on vectors 
revenue <- c(100, 80, 120, 150)
cost <- c(30, 10, 20, 40)
profit <- revenue - cost

print(profit)


prices <- c(80, 100, 40, 160)
prices_inflation <- prices * 1.25
print(prices)
print(prices_inflation)


#what happens when you mix numeric and character values in a vector?
confused_vector <- c(5, 2, "summer", "winter")
print(confused_vector)

#the numeric values turn into character


```

### Some Exercises 2

```{r}

#Task 1
#think about 5 cities you visited or want to visit
#write their names in a vector


#try to guess the temperature now in each city
#write these temperatures in a vector


#multiply these temperatures by 2 then save it in a new vector


#create a logical vector that takes the value TRUE if humans can survive this new temperature and FALSE if they can't










#Task 2
#In the European Social Survey (ESS), the left-right political ideology scale is from 0 to 10, 0 being extreme left and 10 being extreme right
#guess the ideological placement of the following voters and create a vector of the values:
#Christian Democrat, the Left party, Liberal party, Green Party, AFD, and Social Democrat



#modify your vector so the ideological placement would be scaled from 0 to 1





#Task 3
#In the European Social Survey, political interest is scaled from 1-4, where 1 means very interested in politics, and 4 means very uninterested in politics. This is counter intuitive. create a vector for the political interest values of the following survey participants **then** reverse the scale so 4 would mean very interested in politics

#participant 1: very interested
#participant 2: somewhat uninterested
#participant 3: very uninterested
#participant 4: somewhat interested
#participant 5: somewhat uninterested
#participant 6: very interested


#hint: use the formula: new_value = |old_value -(max(old_value) + min (old_value))|









```

Before we move on from vectors, let's look at creating a vector from a sequence of numbers and selecting specific elements from vectors.

```{r}

#fast ways to create a sequence of numbers 
first_ten_numbers <- c(1:10)
print(first_ten_numbers)

first_ten_numbers <- seq(from = 1, to = 10, by = 1)
print(first_ten_numbers)

first_ten_numbers <- seq(1, 10, 1) #because we know the order of the arguments
print(first_ten_numbers)

first_odd_numbers <- seq(1, 10, 2) 
print(first_odd_numbers)

first_even_numbers <- seq(2, 10, 2)
print(first_even_numbers)


#selecting elements from a vector
#let's create a vector with the English alphabet

alphabet <- c("A", "B", "C", "D", "E")

#you can select an element from a vector by providing
#the index (location) of the element in the vector between []
first_letter <- alphabet[1]
print(first_letter)

second_letter <- alphabet[2]
print(second_letter)

#same for the rest of the elements

#remember the length() command, it lets find the last element
length(alphabet) #gives number of elements in the vector
last_letter <- alphabet[length(alphabet)]

#selecting multiple elements
first_three_letters <- alphabet[1:3] #1:3 is similar to c(1,2,3)

print(first_three_letters) #we created a subset of our vector

first_three_letters <- alphabet[c(1,2,3)]

print(first_three_letters)

#how about selecting everything except the first element
all_but_first <- alphabet[2:length(alphabet)]
print(all_but_first)

#even better
all_but_first <- alphabet[-1]
print(all_but_first)

#negative indices means drop that element


#logical values can be used to subset vectors
first_three_letters <- alphabet[c(T,T,T,F,F)]

#TRUE means to include this element, FALSE means to remove it
#we will explore the true potential of this after some exercises


```

let's take a couple of minutes to play around with sequencing and sub-setting the vectors.

```{r}

#create sequence for the numbers 1 to 7


#create a sequence starting with 10, ending with 100, with increments of 10



#you have a vector of ISO names of countries 
country <- c("DE", "DK", "ES", "FI", "FR", "GB", "IT")


#select Germany (DE)


#Select France "FR"


#Select the first 4 countries


#Select all except Italy (IT)



#Select all except France (FR)


#try to repeat some of the above exercises using a logical vector for selecting



```

We mentioned that you can select from/subset vectors using logical vectors.
But we do not need to manually specify which elements to include or exclude, we can define certain *conditions* for including elements.
For example, if you have a vector of student grades, you can choose to include only students above or below a certain grade.
To define these conditions, we first need to get to know some logical operators.

### Logical Operators

You can use R also for TRUE/FALSE statements for logical statements via the comparison operators

-   `>` greater than
-   `<` smaller than
-   `==` equal
-   `!=` not equal
-   `>=` greater than or equal to
-   `<=` less than or equal to
-   `%in%` equal to at least one element in a group

```{r}


1 < 3 

5 >= 8 

11 != 10  

22 == 22  

7 < 3  

5 <= 2+3 

5 %in% c(2,4,5,7,8)

```

You can also use logical operators:

-   `&` element-wise AND operator. It returns TRUE if both elements are true
-   `|` element-wise OR operator. It returns TRUE if one of the statements is TRUE
-   `!` Logical NOT operator. It returns FALSE if statement is TRUE

```{r}

5 & 4 < 8 

#equivalent to
5 < 8 & 4 < 8


5 | 4 < 8 

#equivalent to
5 < 8 | 4 < 8 


!5 > 2 

#try some of these conditions in the console


```

Now let's use this conditions we learnt for subsetting vectors.

```{r}

#we have a vector for temperatures
temp <- c(20, 41, 10, -2 , 35, -3, 8, 18)

temp == 20 #find room temperature
room_temp <- temp[temp == 20]
print(room_temp)


temp > 30 #find very hot temperatures
hot_temp <- temp[temp > 30]
print(hot_temp)


temp < 0 #find freezing temperatures
freezing_temp <- temp[temp < 0]
print(freezing_temp)


temp > 0 & temp < 15 #find cold temperatures
cold_temp <- temp[temp > 0 & temp < 15 ]
print(cold_temp)


temp > 15 & temp < 30 #find nice temperatures
nice_temp <- temp[temp > 15 & temp < 30]
print(nice_temp)



```

### Some Exercises 3

**Rising populist threat in Germany:**

In a potential not so far future, left and right wing populists in Germany are gaining support across the country.
Germany has about 300 electoral districts, we generate fictional data for these 300 electoral districts.

The goal of the task is to select and count the following districts:

Safe democratic: vote populist \< 25%

significant populist minority: 25% \<= vote populist \< 33%

blocking populist minority: 33% \<= vote populist \< 50% ( can block decisions requiring two-thirds majority)

absolute populist majority: 50% \<= vote populist \< 63%

super populist majority: vote populist \>= 63%

**hint**: you can count the number of elements in a vector with the `length()` function

```{r}

set.seed(2024) 
vote_populist <- rnorm(300, 40, 12) 
#we generate 300 normally distributed random values using rnorm()
#set.seed() makes our random draws reproducible










```

### Matrices

Imagine if you get a vector of student grades, but without student IDs or attendance or majors, how useful is the grades vector then?
That is why we always work with multiple vectors of data bundled together.
In R, one way is to bundle a group of vectors into a **matrix**.

```{r}

#lets create some vectors

student_id <- c(100, 101, 102, 103, 104, 105)
grade <- c(1.3, 2.3, 3.3, 1.7, 2, 3.7)
missed_classes <- c(1, 3, 4, 2, 2, 5)

#you can bind vectors together by rows or by columns
#lefts bind by columns using cbind()
student_data_col <- cbind(student_id, grade, missed_classes)
print(student_data_col)


#lefts bind by rows using rbind()
student_data_row <- rbind(student_id, grade, missed_classes)
print(student_data_row)

#the difference is clearer using the dim()
#the commands tells the size of the matrix
dim(student_data_col)
dim(student_data_row)

#the first dimension is always #rows and second is #columns

#what about the length() command?
length(student_data_col)
length(student_data_row)


#A more direct and less common way of creating a matrix
matrix_example <- matrix(1:20, nrow = 4, ncol = 5, byrow = T) # create numeric matrix
print(matrix_example)

dim(matrix_example) #the dimensions we defined
nrow(matrix_example) #number of rows
ncol(matrix_example) #number of columns


```

### Data Frames

Data Frames are more useful than matrices, since they are able to contain objects with different data types:

```{r}

#you can create a dataframe from a matrix
converted_df <- as.data.frame(matrix_example)

#or by defining vectors (with names)
df_example <-  data.frame(
    country = c("Austria", "England", "Brazil", "Germany"), 
    capital = c("Vienna", "London", "Brasília", "Berlin"), 
    population = c(9, 56, 215, 84),
    visited = c(TRUE, FALSE, FALSE, TRUE)
  )

print(df_example)

#another cool way to inspect a dataframe in RStudio
View(df_example)


#we can select a specific column (variable)
print(df_example[1])
print(df_example["country"])
print(df_example$country)
print(df_example[,1]) #remember [row,column]

#there is a subtle difference to note here
#subsetting = creating a smaller (subset) dataframe
class(df_example[1])           
class(df_example["country"])

#extracting = creating a vector
class(df_example$country)
class(df_example[,1])
class(df_example[[1]])


#selecting a row
print(df_example[1,])

#what we specify both a row AND a column
print(df_example[1,1])


#you can easily create a new column in dataframe
df_example$continent <- c("Europe", "Europe", "South America", "Europe")

df_example$pop_thousand <- df_example$population*1000


View(df_example)


```

### Some Exercises 4

Think of five universities, including the university of Mannheim.
Create a dataframe containing variables for the names of the universities, the number of students attending each year (just guess), which city the universities are in, and the number of people living in each city (again just guess), and whether you applied for that university or not (logical vector).

```{r}

#create the dataframe



#create a new variable for the ratio of students to total people living in the city


#select the column for cities and save it in a vector



#select the row for Uni Mannheim and save it in a vector



#select all columns except the application to university and save it in a new dataframe





```

## The if statements and `ifelse()` function:

This is the logic of a simple **if statement**.
If a test expression is met, then the code within the if statement (body expression) should be executed.

```{r}

grade <- 2.3 

if(grade < 3) {
  print("Good Job")
} 

#the condition after if between the ( ) is met, that is why following code between the { } is executed

grade <- 3.3 

if(grade < 3) {
  print("Good Job")
} 

#Since the condition is not met, nothing happens

#what if, instead of doing nothing when the condition is not met, we want to run another code in that case

grade <- 3.3

if (grade < 3) {
  print("Good Job")
} else {
  print("Life goes on")
} 



grade <- 1.3 

if (grade < 3) {
  print("Good Job")
} else {
  print("Life goes on")
}


#We can write the above if-else statement using a more concise way in R: 
ifelse(grade < 3, "Good Job", "Life goes on")



```

Of course not all our conditions are based on a dichotomous operation, so what we need are if else commands with multiple conditions:

```{r}

grade <- 1.0 

if (grade == 1.0) {
  print("Perfect") 
} else if (grade < 2.0) {
  print("Amazing")
} else if (grade < 3.0) {
  print("Good Job")
} else {
  print("Life goes on")
}

#You just need to add another if function into the body expression of the first else body expression: 

# if (test expression) { body expression } else if(test expression 2) { body expression if} etc...

  
grade <- 2.7

if (grade == 1.0) {
  print("Perfect") 
} else if (grade < 2.0) {
  print("Amazing")
} else if (grade < 3.0) {
  print("Good Job")
} else {
  print("Life goes on")
}

grade <- 3.3

if (grade == 1.0) {
  print("Perfect") 
} else if (grade < 2.0) {
  print("Amazing")
} else if (grade < 3.0) {
  print("Good Job")
} else {
  print("Life goes on")
}


#Again we can do so-called nested ifelse function, using the ifelse command

grade <- 1.7

ifelse(grade == 1.0, "Perfect", 
       ifelse(grade < 2, "Amazing", 
              ifelse(grade < 3, "Good Job", "Life goes on")))

grade <- 3.3 

ifelse(grade == 1.0, "Perfect", 
       ifelse(grade < 2, "Amazing", 
              ifelse(grade < 3, "Good Job", "Life goes on")))

```

How can if statements be useful in real research situations?
let's have a look with a mini-exercise.

In the ESS, ideological placement is coded from 0-10, where 0 is extreme left, and 10 is extreme right.
In your analysis, you only care if someone is left, center, or right.
Create a new variable called pol_cat that takes the values "left" for placement 0-3, "center" for placement 4-6, and "right" for placement 7-10.

Also in the ESS, the answer to the question about belonging to a specific religion is coded 1 = yes, and 2 = no.
Recode this variable so 0 = no and 1 = yes.

```{r}

ess_df <- data.frame(left_right = c(2,5,3,7,1,8,10,0),
                      relg_belog = c(1,2,1,2,2,1,1,2))






```

## End of Session Exercises:

### Exercise 1:

Create a vector called `my_vector` with the values 1,2,3 and check is class.

```{r, eval=FALSE}
#create the vector 
my_vector <- 

#check the class

```

### Exercise 2:

Create a Matrix called `student`.
This should contain information about the `name`, `age` and `major`.
Make three vectors with three entries and bind them together to a the matrix `student`.
Print the matrix.

```{r, eval = FALSE}
#Create the vectors 

name <- 
age <- 
major <- 

#Create the matrix
student <- 
  
#Print the matrix

```

### Exercise 3:

Write an if-else statement that checks if a given number is positive or negative.
If the number is positive, print "Number is positive", otherwise print "Number is negative".
Feel free to decide if you want to use the ifelse function or the ifelse condition.

```{r}

#Assigning the number to the object "number"
number <- -4



```

### Exercise 4:

Write an if-else ladder that categorizes a student's grade based on their score.
The grading criteria are as follows:

Score \>= 90: "A" Score \>= 80 and \< 90: "B" Score \>= 70 and \< 80: "C" Score \>= 60 and \< 70: "D" Score \< 60: "F"

```{r}

```
