---
title: "Intro to R. Session 3: Solutions"
output:
  html_notebook:
    toc: TRUE
    toc_level: 3
    code_folding: "none"
editor_options: 
  markdown: 
    wrap: sentence
---

## Exercises Chapter 5: 

### Exercise 1: Write a `for` loop that prints the square of each number from 1 to 10

```{r}
#Assigning an object for a better workflow
number <- 10

#The Loop 
for (i in 1:number) {
  sq <- i^2
  print(sq)
}

```


### Exercise 2: Do the same, but this time defining a function that squares numbers.

```{r}
#Defining a function for squaring

sq <- function (x) {
  
  result <- x^2
  print(result)
  
}

#Defining a vector containing a vector from 1 to 10 
numbers <- c(1:10) 

#Applying the number 
sq(numbers)

```


### Exercise 3: This is the midnight formula separated in two equations:

$$

x_{1,2} = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}

$$

Make **one function** for the midnight formula, so the output are both. Test it with a = 2, b = -6, c = -8

**Hint**: You need two split up the formula into two equations with two outputs.

```{r}
mnf <- function (a,b,c) {
  
  x_1 <- (-b + sqrt(b^2 - 4*a*c))/(2*a)
  
  print(x_1)
  
  x_2 <- (-b - sqrt(b^2 - 4*a*c))/(2*a)
  
  print(x_2)
}


mnf(2, -6, -8)

```


### Exercise 4: Zodiac Fortune Teller 

In the last exercise of this course, we will learn about our future based on our Zodiac Sign. For this we well need a `function`, a `loop` and an `ifelse` condition or function. 

a. First we define a function called `zodiacFortuneTeller()`. This function takes only your sign as input 

b. Afterwards we need to define our fortunes for each zodiac with a vector 

c. Then we need to define a vector with contains the zodiac signs

d. Lastly we need a `for` loop, which returns the respective fortune for each zodiac sign. 

e. This fortune should be returned 

```{r}

fortune <- function(sign) {
  
  #Fortunes vector 
  fortunes <- c("Exciting opportunities are on the horizon for you!",
              "Your creative talents will shine brightly this month!",
              "Financial abundance is heading your way!",
              "You will experience a period of personal growth and
              self-discovery!",
              "New friendships will bring joy and fulfillment into you
              life!",
              "Your hard work and dedication will pay off in unexpected
              ways!",
              "Travel plans will bring you exciting adventures and memorable
              experiences!",
              "A long-awaited dream will finally become a reality!",
              "Your positive attitude and optimism will attract great
              success!",
              "Love and romance will blossom in your life!",
              "Embrace change and seize new opportunities with confidence!",
              "Your intuitive insights will guide you to make wise
              decisions!",
              "A new career opportunity will open doors to success and
              fulfillment!",
              "You have the power to overcome any challenges that come your
              way!",
              "Your generosity and kindness will be rewarded abundantly!",
              "Self-care and self-love will lead to inner peace and harmony!"
              )
  
  #Zodiac signs vector 
  zodiac_signs <- c("Aries", "Taurus", "Gemini", "Cancer", "Leo", "Virgo",
                    "Libra", "Scorpio", "Sagittarius", "Capricorn",
                    "Aquarius", "Pisces")
  #For loop 
  for (i in 1:length(zodiac_signs)) {
    if (sign == zodiac_signs[i]) {
      return(fortunes[i])
    }
  }
} 

# Let us see our fortune 

fortune("Aries") #Do not forget to put your sign in quotation marks

```

