---
title: "Intro to R. Session 3: Solutions"
output:
  html_notebook:
    toc: TRUE
    toc_level: 3
    code_folding: "none"
editor_options: 
  markdown: 
    wrap: sentence
---

## Exercises Chapter 3:

### Exercise 1: Power to the People

```{r}
summary(comp$populism)
mean(comp$populism)
median(comp$populism)
quantile(comp$populism, na.rm = T)
quantile(comp$populism, 0.75, na.rm = T) - quantile(comp$populism, 0.25, na.rm = T)
hist(comp$populism, breaks = 40)
```

### Exercise 2: Anti-Immigration Sentiment, That's what `the People` want...?

```{r}
cor(comp$generalwill, comp$immigration, use = "complete.obs") 
cor(comp$generalwill, comp$immigration, method = "spearman", use = "complete.obs")
cor(comp$generalwill, comp$immigration, method = "kendall", use = "complete.obs")
```

All three correlation coefficients indicate a medium sized negative correlation. Judging from the codebook, this indicates that the more a party is in favor of immigration, the less it claims to represent a `general will` within the society, a populist concept and one of populism's main dimensions. 
